%!TEX root = ../dissertation.tex
% Abstract

Since it's first incarnation in the 70s under the "NAVSTAR" name, satellite
positioning acquired more and more relevance in the life of everyone. Directly
or indirectly, we all rely on technological aids to precisely define our
position. This exponential trend uncovered in recent years the need to secure
the information transmitted by legitimate satellites: military signals are
encrypted before transmission with schemes that allow only an authorized and
restricted set of users, but for the civil case such a restriction clashes with
the idea of having an open signal accessible by anyone. Yet, leaving the
transmitted data unprotected opens the door to various kinds of attack. To name
a clear example, it suffices to think about spoofing the signal received by an
aircraft, in a way that modifies its trajectory to achieve catastrophic
consequences.

\vspace{\baselineskip}

It is with the introduction of a satellite constellation operated by the
European Union under the name of Galileo that the topic of data authentication
for open GNSS signal took a practical turn. Some authentication schemes have
been developed on paper to suit the low data rate available, resulting in ESA
releasing an update to Galileo's ICD to include an implementation of TESLA, a
protocol suited to authenticate streams of data that makes no assumption on the
time when a receiver starts receiving the radio signal.

In this paper a critical analysis of the proposed specification is performed, to
understand the implications on the receiver side in terms of performance and
resource consumption.
