%!TEX root = ../dissertation.tex
% the acknowledgments section

I wish to thank a few amazing individuals that guided and supported me through
the challenge of writing a master thesis in the midst of a job change, a
newborn relationship, and a move. The first is Gianluca Caparra, who showed
admirable patience and dedication in holding my back. Gianluca, I'm grateful for
life for you always finding time to answer my questions, jump into skype calls,
and review my early drafts. I truly couldn't have made it without you.

Second, I would like to thank Nicola Laurenti for having given me the freedom of
doing my research from abroad, without a trace of micromanagement and without
questioning my steps (even after dropping the ball for two years and showing up
for the decisive retry). Nicola, I'm extremely grateful for you having made the
process easier and more enjoyable.

I also want to thank my parents, who also never questioned my choices but rather
gave me the credit and trust I needed even if it was hard to see where I was
going. At the same time, I'm grateful to them because they never gave up but
kept the thought of this degree in my head for so many years. Giorgio and
Lorena, I owe my resilience (even if sometimes it borders with stubborness) to
you.

In addition, my gratefulness to those friends, near and far, that inspired and
moved me over these years. In particular, Nikos and Mari: I'm grateful and
honored to have you as friends, and grateful for the challenges and tough
conversations we went through together.

Finally, but most importantly, I want to thank the star that since some time is
brightening my path. Sarah, you're the most wonderful partner anyone could ever
desire. If I was to mention all the things I'm grateful for, the pages of this
work won't suffice to contain them: for one, I truly couldn't have gone so far
so fast without your unconditional support, and I'm grateful for you having
walked the hard path with me. Even as I write these words I can't help feeling
moved at the miracle of us. Thank you for being as you are, you made true love
a reality in my life.
